Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-23 MS_thesis.tex
Add a cleardoublepage after the new appendix to make sure headers work out
Minor rearrangement
Make tracked changes in response to reviewer comments

	This thesis is dedicated to obeying the inscrutable exhortations of your soul.
	\extrafootertext{\tiny Calvin and Hobbes is the property of Mr. Bill Watterson and Andrews McMeel Syndication. This comic has been reproduced with explicit permission from Andrews McMeel Syndication.}
	}

@@

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, much of this foundation has been built using models of infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in a world where population sizes  routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and may reverse the direction of evolution predicted by infinite-population frameworks. My SDEs also recover well-known results such as the replicator-mutator equation and the Price equation in the infinite population limit, and help explain and organize several previous studies under a single theoretical framework. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations. I also discuss the implications for designing evolutionary agent-based models and empirical studies.
}

@@

I am also grateful to various non-human companions, canine, feline, and more exotic. The dogs and cats of ISI are more helpful to my mental health than I could possibly articulate. Even though my thesis work may seem dreadfully abstract, running after pretty animals in the Western Ghats and other wild places of India is how I even got interested in science in the first place, and I certainly wouldn't be here today if the animals weren't so fascinating :)

@@

%Import the Appendices
\chapter{From It\^{o} to Fokker-Planck}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}
\chapter{Some Examples}\label{App_examples}
\input{ind_files/99_Appendix_D_examples}
\chapter{A more elegant representation of the stochastic integrals of Chapter \ref{chap_unification}}\label{App_alternative_representation}
\input{ind_files/99_Appendix_E_lemma}

@@

%Ask the bibliography headers to say 'References' 
\fancyhead[CO]{\rule[-4ex]{0pt}{4ex}\footnotesize\itshape References}
%Add the phantomsection so that hyperref knows where to go
\cleardoublepage
\phantomsection
\vspace*{1.3cm}
%Add a title for the bibliography
{\noindent\fontsize{48}{60}\selectfont \textbf{References}}
%Add an entry in the table of contents
\addcontentsline{toc}{part}{References}
%Change the font for the bibliography
\renewcommand*{\bibfont}{\footnotesize}	
%Typeset the bibliography little differently using ragged2e
\begingroup
\RaggedRight
\vspace{5ex}
%Add the bibliography
\printbibliography[heading=none]
\endgroup
%And we're done :)
\end{document}

>>>>

	This thesis is dedicated to obeying the inscrutable exhortations of one's soul.
	\extrafootertext{\tiny Calvin and Hobbes is the property of Mr. Bill Watterson and Andrews McMeel Syndication. This comic has been reproduced with explicit permission from Andrews McMeel Syndication.}
	}

@@

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
\textcolor{red}{\st{Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, much of this foundation has been built using models of infinite populations, ignoring the effects of demographic stochasticity.}} Finite population models in standard evolutionary theory usually assume a fixed population size and are of limited applicability in a world where population sizes routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs generically reveal a directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and may reverse the direction of evolution predicted by infinite-population frameworks. \textcolor{blue}{The strength of noise-induced selection depends on the difference in turnover rates between types and the total population size, and my very general derivation helps unify and organize several previous studies under a single set of equations.} My SDEs also recover well-known results such as the replicator-mutator equation, the Price equation, \textcolor{blue}{and Fisher's fundamental theorem} in the infinite population limit, \textcolor{blue}{illustrating consistency with known formal descriptions of evolution}. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields equations such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit \textcolor{blue}{and provides an alternative approach to modeling the evolution of quantitative traits that is more accessible than current measure-theoretic approaches.}. \textcolor{red}{\st{My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations.}}I also discuss the implications \textcolor{blue}{of noise-induced selection} for designing evolutionary agent-based models and empirical studies.
}

@@

I am also grateful to various non-human companions, canine, feline, and more exotic. The dogs and cats of \textcolor{blue}{my home, the Indian Statistical Institute} \textcolor{blue}{(}ISI\textcolor{blue}{) Bangalore,} are more helpful to my mental health than I could possibly articulate. Even though my thesis work may seem dreadfully abstract, running after pretty animals in the Western Ghats and other wild places of India is how I even got interested in science in the first place, and I certainly wouldn't be here today if the animals weren't so fascinating :)

@@

%Import the Appendices
\chapter{From It\^{o} to Fokker-Planck}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}
\chapter{Some Examples}\label{App_examples}
\input{ind_files/99_Appendix_D_examples}
\chapter{A more elegant representation of the stochastic integrals of Chapter \ref{chap_unification}}\label{App_alternative_representation}
\input{ind_files/99_Appendix_E_lemma}
\chapter{Behavior of our fundamental equation for trait frequencies at boundaries}\label{App_well_behaved}
\input{ind_files/99_Appendix_F_well_behaved}

@@

%Ask the bibliography headers to say 'References' 
\cleardoublepage
\fancyhead[CO]{\rule[-4ex]{0pt}{4ex}\footnotesize\itshape References}
%Add the phantomsection so that hyperref knows where to go
\cleardoublepage
\phantomsection
\vspace*{1.3cm}
%Add a title for the bibliography
{\noindent\fontsize{48}{60}\selectfont \textbf{References}}
%Add an entry in the table of contents
\addcontentsline{toc}{part}{References}
%Change the font for the bibliography
\renewcommand*{\bibfont}{\footnotesize}	
%Typeset the bibliography little differently using ragged2e
\begingroup
\RaggedRight
\vspace{5ex}
%Add the bibliography
\printbibliography[heading=none]
\endgroup
%And we're done :)
\end{document}

### 2023-04-23 ind_files/01_Introduction.tex
Make tracked changes in response to reviewer comments

A general approach need not (and often will not) be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”~\citep{kingsland_modeling_1985}. MacArthur was speaking primarily about biological generalities and special cases, but in a related vein, if our language of choice for expressing our general events is mathematics, making any non-trivial observations in complex fields such as ecology and evolution often requires \emph{mathematical} approximations and idealizations that may exclude or underplay several `low-level' model-specific details in favor of a more general description at a `higher' level achieved in some limit that only contains a small number of `model-independent' quantities, often derived from first principles. Formulating and studying such general frameworks can be greatly beneficial as an aid to thinking, sometimes precisely \emph{because} `blurry eyed' thinking that begins from a small set of fundamental first principles and only looks for general broad-brush regularities can be much more insightful than accounting for every little detail or special case. The success of such an approach is perhaps best illustrated by the success of statistical mechanics in physics - statistical mechanics was essentially born from the idea that various useful statements about systems with many moving parts can be made without the need for knowing the excruciating details of every single moving part, and indeed, starting from first principles, this sort of explicitly `blurry-eyed' thinking that only looked at approximate properties was shown to be able to recover the phenomenological laws of thermodynamics as \emph{statistical} laws.

@@

One of the general guiding principles of much of this mathematization has been the assumption that incorporating the reality of finite population sizes into models leads to no major qualitative differences in behavior, only `adding noise' or `blurring out' the predictions of simpler infinite population models~\citep{page_unifying_2002}. Consequently, several major theoretical frameworks in the field, such as adaptive dynamics, are explicitly formulated in deterministic terms at the infinite population size limit. However, this assumption is largely unjustified, and since populations in the real world are finite and stochastic, checking whether stochastic models differ from their deterministic analogs is vital to furthering our understanding of the fundamentals of population biology~\citep{hastings_transients_2004, coulson_skeletons_2004, shoemaker_integrating_2020}. Today, we increasingly recognize that incorporating the finite and stochastic nature of the real world routinely has much stronger consequences than simply `adding noise' to deterministic expectations~\citep{boettiger_noise_2018}, with important consequences for both ecological~\citep{schreiber_does_2022} and evolutionary~\citep{delong_stochasticity_2023} theory. In ecology and evolution, stochastic models need not exhibit phenomena predicted by their deterministic analogues~\citep{proulx_what_2005, johansson_will_2006, claessen_delayed_2007,  wakano_evolutionary_2013, debarre_evolutionary_2016, johnson_two-dimensional_2021}. In addition, they exhibit novel phenomena not predicted by the deterministic approximations~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015, veller_drift-induced_2017, delong_stochasticity_2023}, for example, even completely `reversing' the predictions of deterministic models~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,constable_demographic_2016,mcleod_social_2019}.

@@

Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton'~\citep{coulson_skeletons_2004} in an ad-hoc fashion, can very easily result in models that are not self-consistent~\citep{strang_how_2019}. Further, this procedure of adding noise in an ad-hoc manner provides no insight into the mechanistic factors actually responsible for the stochasticity in the first place. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are self-consistent, much more natural, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model~\citep{black_stochastic_2012,strang_how_2019}. Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons~\citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography'~\citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics~\citep{doebeli_towards_2017}. In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. The mathematical formalism itself is very general and applies equally to the `high level' forces of population genetics and the `high level' forces of community ecology as postulated by~\citet{vellend_theory_2016}, though I will mostly stick to the population genetics interpretation in my discussions.

>>>>

A general approach need not (and often will not) be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”~\citep{kingsland_modeling_1985}. MacArthur was speaking primarily about biological generalities and special cases, but in a related vein, if our language of choice for expressing our general events is mathematics, making any non-trivial observations in complex fields such as ecology and evolution often requires \emph{mathematical} \textcolor{blue}{`coarse-graining'} \textcolor{red}{\st{approximations and idealizations}} that \textcolor{red}{\st{may exclude or underplay}} ignores or `lumps together' several `low-level' model-specific details in favor of a more general description at a `higher' level achieved in some limit that only contains a small number of `model-independent' quantities, often derived from first principles. Formulating and studying such general frameworks can be greatly beneficial as an aid to thinking, sometimes precisely \emph{because} `blurry eyed' thinking that begins from a small set of fundamental first principles and only looks for general broad-brush regularities can be much more insightful than accounting for every little detail or special case. The success of such an approach is perhaps best illustrated by the success of statistical mechanics in physics - statistical mechanics was essentially born from the idea that various useful statements about systems with many moving parts can be made without the need for knowing the excruciating details of every single moving part, and indeed, starting from first principles, this sort of explicitly `blurry-eyed' thinking that only looked at approximate properties was shown to be able to recover the phenomenological laws of thermodynamics as \emph{statistical} laws.

@@

One of the general guiding principles of much of this mathematization has been the assumption that incorporating the reality of finite population sizes into models leads to no major qualitative differences in behavior, only `adding noise' or `blurring out' the predictions of simpler infinite population models~\citep{page_unifying_2002}. Consequently, several major theoretical frameworks in the field, such as adaptive dynamics, are explicitly formulated in deterministic terms at the infinite population size limit. However, this assumption is largely unjustified, and since populations in the real world are finite and stochastic, checking whether stochastic models differ from their deterministic analogs is vital to furthering our understanding of the fundamentals of population biology~\citep{hastings_transients_2004, coulson_skeletons_2004, shoemaker_integrating_2020}. Today, we increasingly recognize that incorporating the finite and stochastic nature of the real world routinely has much stronger consequences than simply `adding noise' to deterministic expectations~\citep{boettiger_noise_2018}, with important consequences for both ecological~\citep{schreiber_does_2022} and evolutionary~\citep{delong_stochasticity_2023} theory. In ecology and evolution, stochastic models need not exhibit phenomena predicted by their deterministic analogues~\citep{proulx_what_2005, johansson_will_2006, claessen_delayed_2007,  wakano_evolutionary_2013, debarre_evolutionary_2016, johnson_two-dimensional_2021}. In addition, they exhibit novel phenomena not predicted by the deterministic approximations~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015, veller_drift-induced_2017, delong_stochasticity_2023}\textcolor{red}{\st{, for example, even completely ‘reversing’ the predictions of deterministic models~(Houchmandzadeh and Vallade, 2012; Houchmandzadeh, 2015; Constable et al., 2016; McLeod and Day, 2019)}}.

\textcolor{blue}{A striking example of such novel phenomena is given by the complete `reversal' of the predictions of deterministic models that are seen in many finite-population evolutionary models~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,constable_demographic_2016,mcleod_social_2019}. These latter models work with two competing types of individuals (in various model-specific contexts, ex: cheaters and cooperators in evolutionary games), and find in their specific models that while deterministic models in the infinite-population limit predict one particular type being superior due to natural selection, the stochastic finite-population models evolve in the exact opposite direction, with the type that is classically `disfavored' in the infinite-population limit having a higher fixation probability. This phenomenon of selection for the classically disfavored type that leads to the `reversal' of the prediction of deterministic natural selection has been dubbed `noise-induced selection'~\citep{constable_demographic_2016,mcleod_social_2019,week_white_2021}, and has been seen in several particular models in fields as diverse as epidemiology~\citep{kogan_two-strain_2014,humplik_evolutionary_2014,parsons_pathogen_2018,day_price_2020}, cell-cycle dynamics~\citep{wodarz_effect_2017}, life-history evolution~\citep{gillespie_natural_1974,veller_drift-induced_2017,kuosmanen_turnover_2022}, and social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,constable_demographic_2016,mcleod_social_2019,wang_reproductive_2023}. However, despite these suggestions that stochastic finite-population effects can have profound consequences in a wide variety of biological systems, as yet, no general model-independent description of noise-induced selection analogous to the Price equation as a description for classical selection and transmission bias exists in the literature.}

@@

Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton'~\citep{coulson_skeletons_2004} in an ad-hoc fashion, can very easily result in models that are \textcolor{blue}{ill-behaved} \textcolor{red}{\st{not self-consistent}}. \textcolor{blue}{For example, such models may predict negative population sizes, or lead to biologically unreasonable predictions such as `extinction time paradoxes' where, for instance, a strong Allee effect can actually appear to increase time to extinction even when a population begins below the Allee threshold, a clearly unreasonable prediction that vanishes if noise is derived properly from first principles}~\citep{black_stochastic_2012,strang_how_2019}. Further, this procedure of adding noise in an ad-hoc manner provides no insight into the mechanistic factors actually responsible for the stochasticity in the first place. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are self-consistent, much more natural, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model~\citep{black_stochastic_2012,strang_how_2019}. Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons~\citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography'~\citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics~\citep{doebeli_towards_2017}. In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. The mathematical formalism itself is very general and applies equally to the `high level' forces of population genetics and the `high level' forces of community ecology as postulated by~\citet{vellend_theory_2016}, though I will mostly stick to the population genetics interpretation in my discussions.
